---
title: "Analysis_RNAseq_Organoids"
author: "MaikTungsten"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r, include=FALSE}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(reshape2)
library(stringr)
library(ggpubr)
library(readxl)
library(AnnotationDbi)
library(topGO)
library(org.Hs.eg.db)
library(limma)
library(DESeq2)
library(pheatmap)
```

# Overview

The dataset encompasses two cell states - Healthy and Glioma cells - subjected to a diverse set of treatments (control, DMSO, 6 drugs) - each analysed in triplicates.

# 1. Load the countsTable

Load the data. Note: the data has been subjected to prior inspection, cleaning and curation in a Jupyter Notebook.

```{r}
countsLOG_in <- as.data.frame(read_csv("../post_processing_countData/Log2_normalized_countData_minus_outliers.csv"))
countsRAW_in <- as.data.frame(read_csv("../post_processing_countData/Raw_countData_minus_outliers.csv"))
countsTPM_in <- as.data.frame(read_csv("../post_processing_countData/TPM_normalized_countData_minus_outliers.csv"))

# Example: TPM data
head(countsTPM_in)
```

```{r}
# Convert Geneid column to rownames
rownames(countsTPM_in) <- countsTPM_in[,1]
countsTPM <- countsTPM_in[,-1]

rownames(countsLOG_in) <- countsLOG_in[,1]
countsLOG <- countsLOG_in[,-1]

rownames(countsRAW_in) <- countsRAW_in[,1]
countsRAW <- countsRAW_in[,-1]

head(countsTPM)
```

#### Load annotation

```{r}
anno <- as.data.frame(read_excel('../post_processing_countData/AW_RNAseq_2025.xlsx', sheet='Final_annotation'))
rownames(anno) <- c(anno$ID)
head(anno)
```


# 2. Inspect the data by PCA

PCA (Principal component analysis) reduces the dimensionality of the dataset to allow for visual inspection. Thereby, we can see the shifts in gene expression between cell types and treatment, for instance.

```{r}
# Define a function for PCA

pca <- function(x){
  
  tpm_pca <- prcomp(t(as.matrix(x)), center = TRUE, scale = TRUE)
  
  # Get contributions of variance of each PC
  pc1 <- summary(tpm_pca)$importance[2,1]*100
  pc2 <- summary(tpm_pca)$importance[2,2]*100
  
  # Configure dataframe for PC1 and PC2 coordinates for all samples
  pcs_1_2 <- tpm_pca$x[,1:2]
  pcs_1_2_sample <- data.frame(cbind(pcs_1_2, rownames(pcs_1_2)))
  colnames(pcs_1_2_sample) <- c("PC1", "PC2", "Sample")
  
  # Transform coordinates to values that can be calculated with
  pcs_1_2_sample$PC1 <- as.numeric(as.character(pcs_1_2_sample$PC1))
  pcs_1_2_sample$PC2 <- as.numeric(as.character(pcs_1_2_sample$PC2))
  
  # Plot data
  p <- ggplot(pcs_1_2_sample, aes(x = PC1, y = PC2, color = Sample))+
    geom_point()+
    theme_minimal()+
    xlab(paste0("PC1 - ", pc1, " % of variance"))+
    ylab(paste0("PC2 - ", pc2, " % of variance"))+
    theme(legend.position = "right")
  
  return(p)
}
```

```{r}
# PCA with all samples, TPM normalised
pca(countsTPM)
```

```{r}
# PCA with all samples, log2 normalized
pca(countsLOG)
```

```{r}
# PCA with all samples, raw counts
pca(countsRAW)
```

#### PCA with sample annotation

```{r}
# Define a function for PCA by cell state

pca_by_state <- function(x, anno){
  
  tpm_pca <- prcomp(t(as.matrix(x)), center = TRUE, scale = TRUE)
  
  # Get contributions of variance of each PC
  pc1 <- summary(tpm_pca)$importance[2,1]*100
  pc2 <- summary(tpm_pca)$importance[2,2]*100
  
  # Configure dataframe for PC1 and PC2 coordinates for all samples
  pcs_1_2 <- tpm_pca$x[,1:2]
  sample_info <- anno[rownames(pcs_1_2),c('Cell_state', 'Treatment', 'Replicate')]
  pcs_1_2_sample <- data.frame(cbind(pcs_1_2, rownames(pcs_1_2), sample_info))
  colnames(pcs_1_2_sample) <- c("PC1", "PC2", "Sample", "Cell_state", "Treatment", "Replicate")
  
  # Transform coordinates to values that can be calculated with
  pcs_1_2_sample$PC1 <- as.numeric(as.character(pcs_1_2_sample$PC1))
  pcs_1_2_sample$PC2 <- as.numeric(as.character(pcs_1_2_sample$PC2))
  
  # Plot data
  p <- ggplot(pcs_1_2_sample, aes(x = PC1, y = PC2, color = Cell_state))+
    geom_point()+
    theme_minimal()+
    xlab(paste0("PC1 - ", pc1, " % of variance"))+
    ylab(paste0("PC2 - ", pc2, " % of variance"))+
    theme(legend.position = "right")
  
  return(p)
}
```

Plot PCA with annotation of samples.

```{r}
pca_by_state(countsTPM, anno)
pca_by_state(countsLOG, anno)
pca_by_state(countsRAW, anno)
```

```{r}
# Define a function for PCA by treatment

pca_by_treatment <- function(x, anno){
  
  tpm_pca <- prcomp(t(as.matrix(x)), center = TRUE, scale = TRUE)
  
  # Get contributions of variance of each PC
  pc1 <- summary(tpm_pca)$importance[2,1]*100
  pc2 <- summary(tpm_pca)$importance[2,2]*100
  
  # Configure dataframe for PC1 and PC2 coordinates for all samples
  pcs_1_2 <- tpm_pca$x[,1:2]
  sample_info <- anno[rownames(pcs_1_2),c('Cell_state', 'Treatment', 'Replicate')]
  pcs_1_2_sample <- data.frame(cbind(pcs_1_2, rownames(pcs_1_2), sample_info))
  colnames(pcs_1_2_sample) <- c("PC1", "PC2", "Sample", "Cell_state", "Treatment", "Replicate")
  
  # Transform coordinates to values that can be calculated with
  pcs_1_2_sample$PC1 <- as.numeric(as.character(pcs_1_2_sample$PC1))
  pcs_1_2_sample$PC2 <- as.numeric(as.character(pcs_1_2_sample$PC2))
  
  # Plot data
  p <- ggplot(pcs_1_2_sample, aes(x = PC1, y = PC2, color = Treatment, shape = Cell_state))+
    geom_point()+
    theme_minimal()+
    xlab(paste0("PC1 - ", pc1, " % of variance"))+
    ylab(paste0("PC2 - ", pc2, " % of variance"))+
    theme(legend.position = "right")
  
  return(p)
}
```

Plot PCA with annotation of samples.

```{r}
pca_by_treatment(countsTPM, anno)
pca_by_treatment(countsLOG, anno)
pca_by_treatment(countsRAW, anno)
```

# 3. Compare samples among each other by Pearson correlation

```{r}
# Pearson Correleation between samples
res <- cor(countsTPM, method = 'pearson')

anno_for_res <- anno[rownames(res), c("Cell_state", "Treatment")]
anno_for_res

pheatmap(res, annotation_row = anno_for_res, annotation_col = anno_for_res)
```


# 4. Plot a heatmap

In a heatmap, one can visualize the overall expression of genes. Here, we filter out the 2,000 most variable genes and z-score-normalize each gene that across all samples.

```{r}
# Define function for filtering top expressing genes and plotting a heatmap

heatmap_plot <- function(x, anno, a="row", f=10){
  library(pheatmap)
  x <- x[order(rowVars(as.matrix(x), useNames = TRUE),decreasing=TRUE),]
  tpm_mat <- as.matrix(x)
  rownames(tpm_mat) <- rownames(x)
  tpm_mat_selected <- tpm_mat[1:2000,]
  #my_hclust_gene <- hclust(dist(tpm_mat_selected), method = "complete")
  #my_gene_col <- as.data.frame(cutree(tree = as.dendrogram(my_hclust_gene), k = f))
  #colnames(my_gene_col) <- "Cluster"
  anno_edit <- anno[colnames(tpm_mat_selected), c("Cell_state", "Treatment")]
  p <- pheatmap(tpm_mat_selected, scale=a, cluster_rows = T, cluster_cols = T, show_rownames = F, annotation_col = anno_edit)
  return(p)
}

#### z-normalize before clustering ####

# Heatmap for all samples with z-score
heatmap_plot(countsTPM, anno, "row", 10)
heatmap_plot(countsLOG, anno, "row", 10)
```

```{r}
# Define function for filtering top expressing genes and plotting a heatmap

heatmap_plot_unscaled <- function(x, anno, f=10){
  library(pheatmap)
  x <- x[order(rowVars(as.matrix(x), useNames = TRUE),decreasing=TRUE),]
  tpm_mat <- as.matrix(x)
  rownames(tpm_mat) <- rownames(x)
  tpm_mat_selected <- tpm_mat[1:2000,]
  #my_hclust_gene <- hclust(dist(tpm_mat_selected), method = "complete")
  #my_gene_col <- as.data.frame(cutree(tree = as.dendrogram(my_hclust_gene), k = f))
  #colnames(my_gene_col) <- "Cluster"
  anno_edit <- anno[colnames(tpm_mat_selected), c("Cell_state", "Treatment")]
  p <- pheatmap(tpm_mat_selected, cluster_rows = T, cluster_cols = T, show_rownames = F, annotation_col = anno_edit)
  return(p)
}

#### z-normalize before clustering ####

# Heatmap for all samples without z-score
heatmap_plot_unscaled(countsTPM, anno, 10)
heatmap_plot_unscaled(countsLOG, anno, 10)
```

One could also extract clusters, which is not necessary at this state of analysis.

```{r}
### Extract genes from distinct clusters
#heatmap_clusters <- function(x, f=6){ # f is number of clusters to extract
#  library(dendextend)
#  x <- x[order(x[,ncol(x)],decreasing=TRUE),]
#  tpm_mat <- as.matrix(x[,3:ncol(x)])
#  rownames(tpm_mat) <- c(x[,1])
#  tpm_mat_selected <- tpm_mat[1:2000,]
#  my_hclust_gene <- hclust(dist(tpm_mat_selected), method = "complete")
#  my_gene_col <- as.data.frame(cutree(tree = as.dendrogram(my_hclust_gene), k = f))
#  return(my_gene_col)
#}

#z <- heatmap_clusters(countsTPM, 12)

```

## 4.1 Gene expression in individual cell types.

Now we take a look at each cell type in individual heatmaps.

```{r}
# Slightly adjusted heatmap plot, selecting for a specified cell type
heatmap_plot <- function(x, anno, a="row", cellType){
  anno_edit <- anno[colnames(x), c("Cell_state", "Treatment", "ID")]
  cell_states <- c(anno_edit[which(anno_edit$Cell_state == cellType), c('ID')])
  df <- x[,cell_states]
  y <- df[order(rowVars(as.matrix(df), useNames = TRUE),decreasing=TRUE),order(colnames(df))]
  tpm_mat <- as.matrix(y)
  rownames(tpm_mat) <- rownames(y)
  tpm_mat_selected <- tpm_mat[1:1000,]
  anno_edit <- anno_edit[colnames(tpm_mat_selected), c("Cell_state", "Treatment")]
  p <- pheatmap(tpm_mat_selected, scale=a, cluster_rows = T, cluster_cols = T, show_rownames = F, annotation_col = anno_edit)
  return(p)
}
```

### 4.1.1 Healthy cells
```{r}
# Healthy cell state, based on TPM
heatmap_plot(countsTPM, anno, "row", cellType = 'Healthy')
# Healthy cell state, based on LOG
heatmap_plot(countsLOG, anno, "row", cellType = 'Healthy')
```
### 4.1.2 Glioma cells
```{r}
# Glioma cell state, based on TPM
heatmap_plot(countsTPM, anno, "row", cellType = 'Glioma')
# Glioma cell state, based on LOG
heatmap_plot(countsLOG, anno, "row", cellType = 'Glioma')
```


# 5. Analysis of cell type and marker-gene expression

```{r}
# Specify cell type and genes of interest
heatmap_plot <- function(x, anno, a="row", cellType, genes){
  anno_edit <- anno[colnames(x), c("Cell_state", "Treatment", "ID")]
  cell_states <- c(anno_edit[which(anno_edit$Cell_state == cellType), c('ID')])
  df <- x[,cell_states]
  y <- df[genes,order(colnames(df))]
  tpm_mat <- as.matrix(y)
  rownames(tpm_mat) <- rownames(y)
  tpm_mat_selected <- tpm_mat
  anno_edit <- anno[colnames(tpm_mat_selected), c("Cell_state", "Treatment")]
  if (a %in% c("row", "col")){
    p <- pheatmap(tpm_mat_selected, scale=a, cluster_rows = T, cluster_cols = F, annotation_col = anno_edit)
  } else {
    p <- pheatmap(tpm_mat_selected, cluster_rows = T, cluster_cols = F, annotation_col = anno_edit)
  }
  
  return(p)
}
```

```{r}
heatmap_plot(countsTPM, anno, "none", cellType = "Healthy", genes = c('Xkr4', 'Rp1', 'Sox17'))
heatmap_plot(countsTPM, anno, "none", cellType = "Glioma", genes = c('Xkr4', 'Rp1', 'Sox17'))
heatmap_plot(countsTPM, anno, "row", cellType = "Healthy", genes = c('Xkr4', 'Rp1', 'Sox17'))
heatmap_plot(countsTPM, anno, "row", cellType = "Glioma", genes = c('Xkr4', 'Rp1', 'Sox17'))
```

Saving plots to pdf.

```{r}
# Scaled rows (z-score)
#pdf("2025-03_final_data_Aylin/Heatmap_CSF_CD4_zscore.pdf")
#heatmap_plot(countsTPM, cellType = "CD4", genes = c('Csf1', 'Csf1r', 'Csf2', 'Csf2ra', 'Csf2rb', 'Csf3'))
#dev.off()
#png("2025-03_final_data_Aylin/Heatmap_CSF_CD4_zscore.png")
#heatmap_plot(countsTPM, cellType = "CD4", genes = c('Csf1', 'Csf1r', 'Csf2', 'Csf2ra', 'Csf2rb', 'Csf3'))
#dev.off()
```


## 5.2 Marker-gene expression plots

```{r}
# Specify a dotplot workflow
plot_markers <- function(x, anno, cellType, genes, scale = "linear"){
  anno_edit <- anno[colnames(x), c("Cell_state", "Treatment", "ID")]
  cell_states <- c(anno_edit[which(anno_edit$Cell_state == cellType), c('ID')])
  
  # Format dataframe
  df <- x[,cell_states]
  y <- df[genes,order(colnames(df))]
  y <- t(y)
  newCols <- append(colnames(y), c('Sample', 'Treatment'))
  y <- cbind.data.frame(y, rownames(y), c(anno_edit$Treatment), stringsAsFactors = FALSE)
  colnames(y) <- newCols
  
  # Melt dataframe
  df_melt <- melt(y, id = c("Sample", "Treatment"))
  colnames(df_melt) <- c('Sample', 'Treatment', 'Gene', 'TPM_normalized_expression')
  
  # Plot data
  if (scale == 'linear'){
      p <- ggplot(df_melt, aes(x = Treatment, y = TPM_normalized_expression, fill = Treatment, colour = Treatment)) +
        geom_dotplot(binaxis='y', stackdir='center', stackratio=1.5, dotsize = 0.6) +
        facet_grid(~ Gene, switch = "x") +
        theme_minimal()+
        theme(panel.spacing.x = grid::unit(0, "mm"),
            strip.placement = "outside",
            strip.background = element_blank(),
            axis.text.x = element_text(angle = 45, hjust = 1))+
        stat_summary(fun.y=mean, geom="point", shape=10,
                     size=3, color="grey")
  } else if (scale == 'log'){
      p <- ggplot(df_melt, aes(x = Treatment, y = TPM_normalized_expression, fill = Treatment, colour = Treatment)) +
        geom_dotplot(binaxis='y', stackdir='center', stackratio=1.5, dotsize = 0.6) +
        facet_grid(~ Gene, switch = "x") +
        theme_minimal()+
        theme(panel.spacing.x = grid::unit(0, "mm"),
            strip.placement = "outside",
            strip.background = element_blank(),
            axis.text.x = element_text(angle = 45, hjust = 1))+
        stat_summary(fun.y=mean, geom="point", shape=10,
                     size=3, color="grey")+
        scale_y_continuous(trans = 'log10')
  }
  

  
  return(p)
}
```

### 5.2.1 Glioma markers
```{r}
# Mouse CD45 = Ptprc
#p <- plot_markers(countsTPM, anno, "Glioma", c("Xkr4", "Sox17", "Rp1"), scale = "linear")
#p
#ggsave("2025-03_final_data_Aylin/Marker_plot_CD4.pdf", width = 30, height = 15, units = "cm")
#ggsave("2025-03_final_data_Aylin/Marker_plot_CD4.png", width = 24, height = 12, units = "cm")
```

```{r}
# Log 10 scale
#p <- plot_markers(countsTPM, anno, "Glioma", c("Xkr4", "Sox17", "Rp1"), scale = "log")
#p
#ggsave("2025-03_final_data_Aylin/Marker_plot_CD4_log.pdf", width = 30, height = 15, units = "cm")
#ggsave("2025-03_final_data_Aylin/Marker_plot_CD4_log.png", width = 24, height = 12, units = "cm")
```


## 5.3 Analysis of specific pathways / GO's

```{r}
# Specify cell type and genes of interest
heatmap_plot <- function(x, anno, a="row", cellType, genes){
  
  anno_edit <- anno[colnames(x), c("Cell_state", "Treatment", "ID")]
  cell_states <- c(anno_edit[which(anno_edit$Cell_state == cellType), c('ID')])
  
  # Format dataframe
  df <- x[,cell_states]
  match_genes <- genes[which(genes %in% rownames(x))]
  y <- df[match_genes,order(colnames(df))]
  tpm_mat <- as.matrix(y)
  rownames(tpm_mat) <- rownames(y)
  tpm_mat_selected <- tpm_mat
  anno_edit <- anno[colnames(tpm_mat_selected), c("Cell_state", "Treatment")]
  p <- pheatmap(tpm_mat_selected, scale=a, cluster_rows = T, cluster_cols = F, annotation_col = anno_edit)
  return(p)
}
```

### 5.3.1 Selected genes from gene sets for CD4 T cells

Synapse pruning and organization
```{r}
# Healthy cells
heatmap_plot(countsTPM, anno, "row", cellType = "Healthy", genes = c('Adgrb3', 'C1ql1', 'Cd47', 'Cdk5', 'Epha4', 'Itgb1', 'Ngef', 'Plxnc1', 'Sarm1', 'Trem2', 'Vangl2'))
heatmap_plot(countsTPM, anno, "none", cellType = "Healthy", genes = c('Adgrb3', 'C1ql1', 'Cd47', 'Cdk5', 'Epha4', 'Itgb1', 'Ngef', 'Plxnc1', 'Sarm1', 'Trem2', 'Vangl2'))
heatmap_plot(countsLOG, anno, "row", cellType = "Healthy", genes = c('Adgrb3', 'C1ql1', 'Cd47', 'Cdk5', 'Epha4', 'Itgb1', 'Ngef', 'Plxnc1', 'Sarm1', 'Trem2', 'Vangl2'))
heatmap_plot(countsLOG, anno, "none", cellType = "Healthy", genes = c('Adgrb3', 'C1ql1', 'Cd47', 'Cdk5', 'Epha4', 'Itgb1', 'Ngef', 'Plxnc1', 'Sarm1', 'Trem2', 'Vangl2'))
```

```{r}
# Glioma cells
heatmap_plot(countsTPM, anno, "row", cellType = "Glioma", genes = c('Adgrb3', 'C1ql1', 'Cd47', 'Cdk5', 'Epha4', 'Itgb1', 'Ngef', 'Plxnc1', 'Sarm1', 'Trem2', 'Vangl2'))
heatmap_plot(countsTPM, anno, "none", cellType = "Glioma", genes = c('Adgrb3', 'C1ql1', 'Cd47', 'Cdk5', 'Epha4', 'Itgb1', 'Ngef', 'Plxnc1', 'Sarm1', 'Trem2', 'Vangl2'))
heatmap_plot(countsLOG, anno, "row", cellType = "Glioma", genes = c('Adgrb3', 'C1ql1', 'Cd47', 'Cdk5', 'Epha4', 'Itgb1', 'Ngef', 'Plxnc1', 'Sarm1', 'Trem2', 'Vangl2'))
heatmap_plot(countsLOG, anno, "none", cellType = "Glioma", genes = c('Adgrb3', 'C1ql1', 'Cd47', 'Cdk5', 'Epha4', 'Itgb1', 'Ngef', 'Plxnc1', 'Sarm1', 'Trem2', 'Vangl2'))
```

Neuroinflammatory response
```{r}
# Healthy cells
heatmap_plot(countsTPM, anno, "row", cellType = "Healthy", genes = c('Cd200', 'Cd200l1', 'Cd200l2', 'Cd200r1', 'Cst7', 'Grn', 'Ifnb1', 'Igf1', 'Ldlr', 'Nr1d1', 'Pparg', 'Syt11', 'Tafa3', 'Tnfrsf1b', 'Trem2'))
heatmap_plot(countsTPM, anno, "none", cellType = "Healthy", genes = c('Cd200', 'Cd200l1', 'Cd200l2', 'Cd200r1', 'Cst7', 'Grn', 'Ifnb1', 'Igf1', 'Ldlr', 'Nr1d1', 'Pparg', 'Syt11', 'Tafa3', 'Tnfrsf1b', 'Trem2'))
heatmap_plot(countsLOG, anno, "row", cellType = "Healthy", genes = c('Cd200', 'Cd200l1', 'Cd200l2', 'Cd200r1', 'Cst7', 'Grn', 'Ifnb1', 'Igf1', 'Ldlr', 'Nr1d1', 'Pparg', 'Syt11', 'Tafa3', 'Tnfrsf1b', 'Trem2'))
heatmap_plot(countsLOG, anno, "none", cellType = "Healthy", genes = c('Cd200', 'Cd200l1', 'Cd200l2', 'Cd200r1', 'Cst7', 'Grn', 'Ifnb1', 'Igf1', 'Ldlr', 'Nr1d1', 'Pparg', 'Syt11', 'Tafa3', 'Tnfrsf1b', 'Trem2'))
```

```{r}
# Glioma cells
heatmap_plot(countsTPM, anno, "row", cellType = "Glioma", genes = c('Cd200', 'Cd200l1', 'Cd200l2', 'Cd200r1', 'Cst7', 'Grn', 'Ifnb1', 'Igf1', 'Ldlr', 'Nr1d1', 'Pparg', 'Syt11', 'Tafa3', 'Tnfrsf1b', 'Trem2'))
heatmap_plot(countsTPM, anno, "none", cellType = "Glioma", genes = c('Cd200', 'Cd200l1', 'Cd200l2', 'Cd200r1', 'Cst7', 'Grn', 'Ifnb1', 'Igf1', 'Ldlr', 'Nr1d1', 'Pparg', 'Syt11', 'Tafa3', 'Tnfrsf1b', 'Trem2'))
heatmap_plot(countsLOG, anno, "row", cellType = "Glioma", genes = c('Cd200', 'Cd200l1', 'Cd200l2', 'Cd200r1', 'Cst7', 'Grn', 'Ifnb1', 'Igf1', 'Ldlr', 'Nr1d1', 'Pparg', 'Syt11', 'Tafa3', 'Tnfrsf1b', 'Trem2'))
heatmap_plot(countsLOG, anno, "none", cellType = "Glioma", genes = c('Cd200', 'Cd200l1', 'Cd200l2', 'Cd200r1', 'Cst7', 'Grn', 'Ifnb1', 'Igf1', 'Ldlr', 'Nr1d1', 'Pparg', 'Syt11', 'Tafa3', 'Tnfrsf1b', 'Trem2'))
```


# 6. DE analysis

DESeq2 for identification of DE genes between conditions (for each cell type, compare treatements with each other). More info given in https://bioconductor.org/packages/3.21/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#countmat.

Import the coldata that annotates the samples (colnames) and make sure they are in same order as in raw counts (only raw counts may be put into DESeq2).

DESeq analysis will be performed for each cell type individually here.

```{r}
coldata <- anno
coldata <- coldata[colnames(countsRAW),]
coldata$Treatment <- factor(coldata$Treatment)
coldata$Cell_state <- factor(coldata$Cell_state)
coldata$Replicate <- factor(coldata$Replicate)
head(coldata)
dim(coldata)
```
Raw counts will be used for DE analysis.

```{r}
head(countsRAW)
```

## 6.1 DE objects for distinct cell states

Function to get results for specific contrasts (comparisons), plot MAplot and get list of DEgenes between conditions.

```{r}
getDEanalysis <- function(dds, comparison, countData, coldata){
  res <- results(dds, contrast = comparison, alpha = 0.05) # significance threshold for adjusted p-value of 5 %
  plotMA(res, ylim = c(-8,8), main = paste0(comparison[2],"_vs_",comparison[3]))
  summary(res)
  print(paste0("Significant DE genes at adjusted p-value of 0.05: ", sum(res$padj < 0.05, na.rm=TRUE)))
  resOrdered <- res[order(res$padj),]
  resSig <- subset(resOrdered, padj < 0.05)
  df <- countData[rownames(resSig),order(colnames(countData))]
  anno <- coldata[colnames(df), c("Cell_state", "Treatment")]
  if ((length(rownames(resSig)) < 30) & (length(rownames(resSig)) > 1)){
    pheatmap(df, cluster_cols = FALSE, scale = "row", cluster_rows = TRUE, annotation_col = anno)
  } else if (length(rownames(resSig)) >= 30) {
    pheatmap(df, cluster_cols = FALSE, scale = "row", cluster_rows = TRUE, show_rownames = FALSE, annotation_col = anno)
  }
  
  return(resSig)
}
```

Function to perform a comprehensive DE and enrichment analysis 

```{r}
getDEanalysisFull <- function(dds, comparison, countData, coldata, ontology = "BP",
                              nTopGO = 15) {

  library(topGO)
  library(org.Mm.eg.db)
  library(AnnotationDbi)
  library(ggplot2)
  library(pheatmap)

  ## ──────────────────────────────────────────────
  ## Differential expression
  ## ──────────────────────────────────────────────
  res <- results(dds, contrast = comparison, alpha = 0.05)
  resOrdered <- res[order(res$padj),]
  resSig <- subset(resOrdered, padj < 0.05)

  message("Significant DE genes (padj < 0.05): ", nrow(resSig))

  ## Volcano plot
  res_df <- as.data.frame(res)
  res_df$gene <- rownames(res_df)
  res_df$neglog10padj <- -log10(res_df$padj)
  res_df$sig <- ifelse(res_df$padj < 0.05, "Significant", "Not significant")

  volcano_plot <- ggplot(res_df, aes(x = log2FoldChange, y = neglog10padj, color = sig)) +
    geom_point(alpha=0.7) +
    scale_color_manual(values = c("grey", "#D55E00")) +
    labs(title="Volcano Plot", x="log2 Fold Change", y="-log10(adjusted p)") +
    theme_bw()

  ## ──────────────────────────────────────────────
  ## Split into over- and under-expressed sets
  ## ──────────────────────────────────────────────
  res_over  <- subset(resSig, log2FoldChange > 0)
  res_under <- subset(resSig, log2FoldChange < 0)

  message("Over-expressed genes: ", nrow(res_over))
  message("Under-expressed genes: ", nrow(res_under))

  ## Helper function to perform topGO and generate plots
  run_topGO <- function(sigGenes, universeGenes, label) {

    if (length(sigGenes) < 5) {
      message("Skipping topGO for ", label, " — not enough significant genes.")
      return(list(results=NULL, bar=NULL, dot=NULL))
    }

    geneList <- factor(as.integer(universeGenes %in% sigGenes))
    names(geneList) <- universeGenes

    geneID2GO <- annFUN.org(
      whichOnto = ontology,
      feasibleGenes = universeGenes,
      mapping = "org.Mm.eg.db",
      ID = "SYMBOL"
    )

    GOdata <- new("topGOdata",
                  ontology = ontology,
                  allGenes = geneList,
                  annot = annFUN.GO2genes,
                  GO2genes = geneID2GO,
                  nodeSize = 5)

    resultFisher <- runTest(GOdata, algorithm="weight01", statistic="fisher")

    goRes <- GenTable(GOdata,
                      Fisher = resultFisher,
                      orderBy = "Fisher",
                      topNodes = length(score(resultFisher)))
    goRes$Fisher <- as.numeric(goRes$Fisher)
    goRes <- goRes[goRes$Fisher < 0.05, ]

    if (nrow(goRes) == 0) {
      message("No GO terms enriched for ", label)
      return(list(results=goRes, bar=NULL, dot=NULL))
    }

    ## Select top N
    topN <- min(nTopGO, nrow(goRes))
    goTop <- goRes[1:topN, ]
    goTop$Term <- factor(goTop$Term, levels = rev(goTop$Term))

    ## Bar plot
    barplot <- ggplot(goTop, aes(x = Term, y = -log10(Fisher))) +
      geom_bar(stat="identity", fill="#0072B2") +
      coord_flip() +
      labs(title=paste("Top", topN, ontology, "GO Terms —", label),
           y="-log10(p-value)", x="") +
      theme_bw()

    ## Dot plot
    dotplot <- ggplot(goTop, aes(x = -log10(Fisher), y = Term)) +
      geom_point(size=4, color="#D55E00") +
      labs(title=paste("GO Enrichment Dotplot —", label),
           x="-log10(p-value)", y="") +
      theme_bw()

    return(list(results=goRes, bar=barplot, dot=dotplot))
  }

  universe <- rownames(res)

  ## Run GO for over- and under-expressed genes
  GO_over  <- run_topGO(rownames(res_over), universe, "Over-expressed")
  GO_under <- run_topGO(rownames(res_under), universe, "Under-expressed")

  ## Return everything
  return(list(
    DEgenes_over   = res_over,
    DEgenes_under  = res_under,
    GOresults_over = GO_over$results,
    GOresults_under = GO_under$results,
    GOplots = list(
      over = list(bar = GO_over$bar, dot = GO_over$dot),
      under = list(bar = GO_under$bar, dot = GO_under$dot)
    ),
    Volcanoplot = volcano_plot
  ))
}

```

Function to subset data to dataframe (tissue and treatments of interest) for creation of DESeq object

```{r}
getDEresults <- function(df, coldata, tissue, treatment){

  df_select <- countsRAW[,which(coldata$Cell_state == "Healthy")]

  coldata_select <- coldata[colnames(df_select),c("Cell_state", "Treatment", "Replicate")]
  coldata_select <- coldata_select %>% filter(Treatment %in% c("DMSO", treatment))

  coldata_select$Treatment <- factor(coldata_select$Treatment, levels = c("DMSO", treatment))
  coldata_select$Cell_state <- factor(coldata_select$Cell_state)
  coldata_select$Replicate <- factor(coldata_select$Replicate)

  df_select <- df_select[,rownames(coldata_select)]

  dds <- DESeqDataSetFromMatrix(countData = df_select,
                              colData = coldata_select,
                              design = ~ Treatment)

  dds <- DESeq(dds) # perform DESeq estimations
  resultsNames(dds) # Result names in DESeq object

  return(list(dds=dds, coldata_select=coldata_select, df_select=df_select))


}
```

## 6.1.1 DESeq analysis for Healthy tissue

```{r}
for (treatment in c("AUF", "TMZ")){

  dds_results <- getDEresults(countsRAW, coldata, "Healthy", treatment)
  contrast <- c("Treatment", treatment, "DMSO")
  sig <- getDEanalysis(dds_results$dds, contrast, dds_results$df_select, dds_results$coldata_select)
  if (nrow(sig) > 0) {
    sig_new <- getDEanalysisFull(dds_results$dds, contrast, dds_results$df_select, dds_results$coldata_select)
    print(sig_new$Volcanoplot)
    print(sig_new$GOplots$over$bar)
    print(sig_new$GOplots$over$dot)
    print(sig_new$GOplots$under$bar)
    print(sig_new$GOplots$under$dot)
  }
  
}
```


## 6.1.2 DESeq analysis for Glioma tissue

```{r}
for (treatment in c("AUF", "TMZ")){

  dds_results <- getDEresults(countsRAW, coldata, "Glioma", treatment)
  contrast <- c("Treatment", treatment, "DMSO")
  sig <- getDEanalysis(dds_results$dds, contrast, dds_results$df_select, dds_results$coldata_select)
  if (nrow(sig) > 0) {
    sig_new <- getDEanalysisFull(dds_results$dds, contrast, dds_results$df_select, dds_results$coldata_select)
    print(sig_new$Volcanoplot)
    print(sig_new$GOplots$over$bar)
    print(sig_new$GOplots$over$dot)
    print(sig_new$GOplots$under$bar)
    print(sig_new$GOplots$under$dot)
  }
  
}
```



## 6.2 Use best/largest DE sets and plot heatmaps ####TODO
Plot heatmaps for a specific cell type using DE genes from another to better identify patterns in the data.

```{r}
refined_heatmap <- function(sig, cellType, countData){
  cols <- grep(cellType, colnames(countData))
  
  df <- countData[rownames(sig),cols]
  df <- df[,order(colnames(df))]
  
  if ((length(rownames(df)) < 30) & (length(rownames(df)) > 1)){
    pheatmap(df, scale = "row", cluster_cols = FALSE, cluster_rows = TRUE)
  } else if (length(rownames(df)) >= 30) {
    pheatmap(df, scale = "row", cluster_cols = FALSE, cluster_rows = TRUE, show_rownames = FALSE)
  }
}

```


## 6.3 Use best/largest DE sets and plot heatmaps
Plot heatmaps for a specific cell type using DE genes from another to better identify patterns in the data.

```{r}
all_DE_heatmap <- function(sig, cellType, countData, scaling){
  cols <- grep(cellType, colnames(countData))
  
  df <- countData[sig,cols]
  df <- df[,order(colnames(df))]
  
  pheatmap(df, scale = scaling, cluster_cols = FALSE, cluster_rows = TRUE, show_rownames = FALSE)
}

```

# 7 Gene set-based analysis

## Selected ssGSEA

```{r}
library(GSVA)
library(GSA)
library(GSEABase)

# Define function for ssGSEA analysis
ssGSEA_select <- function(df, GS, keywords = c()){
  
  # Select specific gene sets based on pattern search
  GSnames <- names(GS)
  GS_selectNames <- c()
  for (keyword in keywords) {
    hitNames <- names(GS[str_detect(GSnames, keyword)])
    GS_selectNames <- append(GS_selectNames, hitNames)
  }
  GS_selectNames <- unique(GS_selectNames)
  GS_filtered <- GS[GS_selectNames]
  
  # Extract expression counts as matrix and perform ssGSEA
  expr_mat <- as.matrix(df)
  #params <- gsvaParam(expr_mat, GS_filtered, maxDiff=FALSE)
  if (length(GS_selectNames) > 0){
    a <- gsva(ssgseaParam(expr_mat, GS_filtered))
    return(a)   
  } else {
    a <- "EMPTY"
    return(a)
  }

}

# Define function for ssGSEA analysis
ssGSEA <- function(df, MyGeneSet){
  
  # Select specific gene sets based on pattern search
  GSnames <- names(GS)
  GS_filtered <- GS
  
  # Extract expression counts as matrix and perform ssGSEA
  expr_mat <- as.matrix(df)
  #params <- gsvaParam(expr_mat, GS_filtered, maxDiff=FALSE)
  if (length(GSnames) > 0){
    a <- gsva(ssgseaParam(expr_mat, GS_filtered))
    return(a)   
  } else {
    a <- "EMPTY"
    return(a)
  }
}
```

```{r}
# Define heatmap function for plotting ssGSEA output
# x = ssGSEA input, n = number of GeneSets per heatmap
ssgsea_heatmap <- function(x, anno, n=25, rownames=TRUE){
  library(pheatmap)
  y <- nrow(x)
  anno <- anno[,c("Treatment", "Cell_state")]
  df <- as.matrix(x)
  p <- list()
  i <- 1
  j <- 1
  while (i<=y){
    if ((y-(i+n-1))>0){
      p[[j]] <- pheatmap(df[i:(i+n-1),], cluster_cols = F, cluster_rows = F, show_rownames = rownames, annotation_col = anno)[[4]]
      i <- i+n
      j <- j+1
    } else if ((i+n) > y){
      p[[j]] <- pheatmap(df[i:y,], cluster_cols = F, cluster_rows = F, show_rownames = rownames, annotation_col = anno)[[4]]
      i <- i+n
      j <- j+1
    }
  }
  return(p)
}
```

#### Molecular Hallmark GeneSets

```{r, results='hide'}
cellType <- "Healthy"
df <- countsTPM
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
MyGeneSet <- "../../Analysis_R/Mouse_Hallmark_GS.symbols.gmt.txt"
# Load GeneSets
GS <- GSA.read.gmt(MyGeneSet)
GS <- setNames(GS$genesets, GS$geneset.names)
GS <- lapply(GS, function(x) x[x!= ""])
```

```{r, fig.margin=TRUE}
#word <- c("SYNAPSE_PRUNING", "SYNAPSE_ORGANIZATION", "MICROGLIAL_CELL_ACTIVATION")
#x <- ssGSEA_select(df, GS, word)
#png("2025-03_final_data_Aylin/Heatmap_BPs_MDM.png")
x <- ssGSEA(df, GS)
ssgsea_heatmap(x, new_anno, rownames = TRUE)
#dev.off()
```

```{r, fig.margin=TRUE}
cellType <- "Glioma"
df <- countsTPM
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
x <- ssGSEA(df, GS)
ssgsea_heatmap(x, new_anno, rownames = TRUE)
```

Also test with LOG normalized data instead of TPM data.

```{r, results='hide'}
cellType <- "Healthy"
df <- countsLOG
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
MyGeneSet <- "../../Analysis_R/Mouse_Hallmark_GS.symbols.gmt.txt"
# Load GeneSets
GS <- GSA.read.gmt(MyGeneSet)
GS <- setNames(GS$genesets, GS$geneset.names)
GS <- lapply(GS, function(x) x[x!= ""])
```

```{r, fig.margin=TRUE}
#word <- c("SYNAPSE_PRUNING", "SYNAPSE_ORGANIZATION", "MICROGLIAL_CELL_ACTIVATION")
#x <- ssGSEA_select(df, GS, word)
#png("2025-03_final_data_Aylin/Heatmap_BPs_MDM.png")
x <- ssGSEA(df, GS)
ssgsea_heatmap(x, new_anno, rownames = TRUE)
#dev.off()
```

```{r, fig.margin=TRUE}
cellType <- "Glioma"
df <- countsLOG
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
x <- ssGSEA(df, GS)
ssgsea_heatmap(x, new_anno, rownames = TRUE)
```

Also test with RAW count data instead of TPM data.

```{r, results='hide'}
cellType <- "Healthy"
df <- countsRAW
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
MyGeneSet <- "../../Analysis_R/Mouse_Hallmark_GS.symbols.gmt.txt"
# Load GeneSets
GS <- GSA.read.gmt(MyGeneSet)
GS <- setNames(GS$genesets, GS$geneset.names)
GS <- lapply(GS, function(x) x[x!= ""])
```

```{r, fig.margin=TRUE}
#word <- c("SYNAPSE_PRUNING", "SYNAPSE_ORGANIZATION", "MICROGLIAL_CELL_ACTIVATION")
#x <- ssGSEA_select(df, GS, word)
#png("2025-03_final_data_Aylin/Heatmap_BPs_MDM.png")
x <- ssGSEA(df, GS)
ssgsea_heatmap(x, new_anno, rownames = TRUE)
#dev.off()
```

```{r, fig.margin=TRUE}
cellType <- "Glioma"
df <- countsRAW
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
x <- ssGSEA(df, GS)
ssgsea_heatmap(x, new_anno, rownames = TRUE)
```

CONCLUSION: LOG and TPM normalization seem to yield similar scores per GS.


#### Molecular Tumor Phenotype GeneSets

```{r, fig.margin=TRUE}
cellType <- "Healthy"
df <- countsTPM
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
MyGeneSet <- "../../Analysis_R/Mouse_tumor_phenotype_GS.symbols.gmt.txt"
# Load GeneSets
GS <- GSA.read.gmt(MyGeneSet)
GS <- setNames(GS$genesets, GS$geneset.names)
GS <- lapply(GS, function(x) x[x!= ""])
x <- ssGSEA(df, GS)
ssgsea_heatmap(x, new_anno, rownames = TRUE)
```

```{r, fig.margin=TRUE}
cellType <- "Glioma"
df <- countsTPM
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
x <- ssGSEA(df, GS)
ssgsea_heatmap(x, new_anno, rownames = TRUE)
```


#### Mouse Positional (chromosome-specific) GeneSets

```{r, fig.margin=TRUE}
cellType <- "Healthy"
df <- countsTPM
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
MyGeneSet <- "../../Analysis_R/Mouse_Positional_GS.symbols.gmt.txt"
# Load GeneSets
GS <- GSA.read.gmt(MyGeneSet)
GS <- setNames(GS$genesets, GS$geneset.names)
GS <- lapply(GS, function(x) x[x!= ""])
x <- ssGSEA(df, GS)
ssgsea_heatmap(x, new_anno, rownames = TRUE)
```

```{r, fig.margin=TRUE}
cellType <- "Glioma"
df <- countsTPM
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
x <- ssGSEA(df, GS)
ssgsea_heatmap(x, new_anno, rownames = TRUE)
```


#### Mouse Cell Type Signature GeneSets

```{r, fig.margin=TRUE}
cellType <- "Healthy"
df <- countsTPM
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
MyGeneSet <- "../../Analysis_R/Mouse_cell_type_signature_GS.symbols.gmt.txt"
# Load GeneSets
GS <- GSA.read.gmt(MyGeneSet)
GS <- setNames(GS$genesets, GS$geneset.names)
GS <- lapply(GS, function(x) x[x!= ""])
x <- ssGSEA(df, GS)
ssgsea_heatmap(x, new_anno, rownames = TRUE)
```

```{r, fig.margin=TRUE}
cellType <- "Glioma"
df <- countsTPM
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
x <- ssGSEA(df, GS)
ssgsea_heatmap(x, new_anno, rownames = TRUE)
```


#### Mouse Cell Type Signature GeneSets

```{r, fig.margin=TRUE}
cellType <- "Healthy"
df <- countsTPM
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
MyGeneSet <- "../../Analysis_R/Mouse_cell_type_signature_GS.symbols.gmt.txt"
# Load GeneSets
GS <- GSA.read.gmt(MyGeneSet)
GS <- setNames(GS$genesets, GS$geneset.names)
GS <- lapply(GS, function(x) x[x!= ""])
x <- ssGSEA(df, GS)
ssgsea_heatmap(x, new_anno, rownames = TRUE)
```

```{r, fig.margin=TRUE}
cellType <- "Glioma"
df <- countsTPM
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
x <- ssGSEA(df, GS)
ssgsea_heatmap(x, new_anno, rownames = TRUE)
```


#### Mouse BioCarta Canonical Pathways GeneSets

```{r, fig.margin=TRUE}
cellType <- "Healthy"
df <- countsTPM
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
MyGeneSet <- "../../Analysis_R/Mouse_BioCarta_canonical_pathways_GS.symbols.gmt.txt"
# Load GeneSets
GS <- GSA.read.gmt(MyGeneSet)
GS <- setNames(GS$genesets, GS$geneset.names)
GS <- lapply(GS, function(x) x[x!= ""])
x <- ssGSEA(df, GS)
#ssgsea_heatmap(x, new_anno, rownames = TRUE)
```

```{r, fig.margin=TRUE}
cellType <- "Glioma"
df <- countsTPM
new_anno <- anno[colnames(df),]
new_anno <- new_anno[which(new_anno$Cell_state == cellType),]
df <- df[,rownames(new_anno)]
df <- df[,order(colnames(df))]
new_anno <- new_anno[colnames(df),]
x <- ssGSEA(df, GS)
#ssgsea_heatmap(x, new_anno, rownames = TRUE)
```


